\Title

### 2013-07-31 supplement_body.tex
first inclusions of revision

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time-dependent selection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_timeselec_deadagain}
\caption{
Time-dependent selection reduces fixation of nonsynonymous SNVs. The figure
compares the fixation probability in the time independent model (na\"ive) to
a model with time dependent selection that mimics  an evolving immune system.
It has been found that virus is typically neutralized by serum from a few months
earlier~\citep{richman_rapid_2003} but not by contemporary serum. We model this
evolving immune system by assuming that escaped variants lose their beneficial
effect with a rate proportional to the frequency of the escaped variant. 
Specifically, the selection effect of the escape mutations is
reset to its fitness cost of $-0.02$ with probability
\[ P_\text{recognized}(t) = c \cdot \nu(t), \] 
per generation, where $c$ is a constant coefficient shown in the legend that
encodes the overall efficiency of the host immune system. With increasing
probability of recognition, the fixation of frequent escape mutants is reduced,
while hitch-hiking of synonymous SNVs is not affected. The precise
shape of $\pfix(\nu)$ depends on the details of the $P_\text{recognized}(t)$, and 
we do not think that the high $\pfix(\nu)$ for $\nu<0.2$ is meaningful.
The other parameters for the shown simulations are
the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
$k_A=0.0014$ per generation.
}
\label{fig:tds}
\end{center}
\end{figure}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Within-epitope competition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_epitopes}
\caption{
Competition between escape SNVs in the same epitope reduces fixation of
nonsynonymous SNVs. The figure compares the fixation probability of models
with one, three, or six mutually exclusive escape mutations within the
same epitope. Within epitope competition results in reduced fixation
probabilities of nonsynonymous changes, whereas the synonymous changes behave 
similarly in all cases. We assume that escape can happen at $n$ sites out of 3
consecutive codons and vary $n$.
The fitness landscape of each epitope includes negative epistatic terms, so that
the joint presence of more than one escape mutation is not any more beneficial
for the virus than a single mutation. Specifically, each site has two alleles,
$\pm 1$, where $-1$ is the ancestral one and $+1$ the derived one; the fitness
coefficient of a $k$-tuple of sites within the epitope is $f_k = (-1)^{k-1}
2^{1-n}\eta_\epsilon $, where $\eta_\epsilon$ is the escape rate of the epitope
drawn from an exponential distribution with mean $\epsilon$ and 
$n$ is the number of competing escapes in the epitope. 
In this evolutionary scenario, many escape SNVs start to sweep on different backgrounds within the viral population, but eventually
compete and only one of them fixes. The other parameters for the shown simulations are
the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
$k_A=0.0014$ per generation.
}
\label{fig:wec}
\end{center}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time-dependent selection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_timeselec_deadagain}
\caption{
Time-dependent selection reduces fixation of nonsynonymous SNVs. The figure
compares the fixation probability in the time independent model (na\"ive) to
a model with time dependent selection that mimics  an evolving immune system.
It has been found that virus is typically neutralized by serum from a few months
earlier~\citep{richman_rapid_2003} but not by contemporary serum. We model this
evolving immune system by assuming that escaped variants lose their beneficial
effect with a rate proportional to the frequency of the escaped variant. 
Specifically, the selection effect of the escape mutations is
reset to its fitness cost of $-0.02$ with probability
\[ P_\text{recognized}(t) = c \cdot \nu(t), \] 
per generation, where $c$ is a constant coefficient shown in the legend that
encodes the overall efficiency of the host immune system. With increasing
probability of recognition, the fixation of frequent escape mutants is reduced,
while hitch-hiking of synonymous SNVs is not affected. The precise
shape of $\pfix(\nu)$ depends on the details of the $P_\text{recognized}(t)$, and 
we do not think that the high $\pfix(\nu)$ for $\nu<0.2$ is meaningful.
The other parameters for the shown simulations are
the following: deleterious effect $s_d = 2 \cdot 10^{-3}$, average escape rate $\epsilon
= 0.032$, fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
$k_A=0.0014$ per generation.
}
\label{fig:tds}
\end{center}
\end{figure}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Within-epitope competition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_epitopes}
\caption{
Competition between escape SNVs in the same epitope reduces fixation of
nonsynonymous SNVs. The figure compares the fixation probability of models
with one, three, or six mutually exclusive escape mutations within the
same epitope. Within epitope competition results in reduced fixation
probabilities of nonsynonymous changes, whereas the synonymous changes behave 
similarly in all cases. We assume that escape can happen at $n$ sites out of 3
consecutive codons and vary $n$.
The fitness landscape of each epitope includes negative epistatic terms, so that
the joint presence of more than one escape mutation is not any more beneficial
for the virus than a single mutation. Specifically, each site has two alleles,
$\pm 1$, where $-1$ is the ancestral one and $+1$ the derived one; the fitness
coefficient of a $k$-tuple of sites within the epitope is $f_k = (-1)^{k-1}
2^{1-n}\eta_\epsilon $, where $\eta_\epsilon$ is the escape rate of the epitope
drawn from an exponential distribution with mean $\epsilon$ and 
$n$ is the number of competing escapes in the epitope. 
In this evolutionary scenario, many escape SNVs start to sweep on different backgrounds within the viral population, but eventually
compete and only one of them fixes. The other parameters for the shown simulations are
the following: deleterious effect $s_d = 2 \cdot 10^{-3}$, average escape rate
$\epsilon = 0.032$, fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
$k_A=0.0014$ per generation.
}
\label{fig:wec}
\end{center}
\end{figure}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Simulations with different parameters}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_syn_gentime_2days.pdf}
\caption{
Results of simulations with alternative parameters.
Panel A) Like \figurename~4B in the main text, but simulations are performed using a
generation time of two days. Two selection coefficient of synonymous sites
doubles in units of 1/generations, hence keeps constants in units of 1/days.
}
\label{fig:sims_altparameters}
\end{center}
\end{figure}

### 2013-07-31 synmut.tex
first inclusions of revision

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\pol}{\textit{pol}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\syndiv}{2}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Quantifying selection against synonymous mutations in HIV-1 \env{} evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{hyperref}
\hypersetup{colorlinks,linkcolor=red,citecolor=blue, pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\linespread{1.5}
\begin{document}
\title{\Title}
\author{\Author\\ \Affiliation}
%\affiliation{\Affiliation}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle

\newpage
%\begin{abstract}
\section*{Abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
or neutralizing antibodies target a new epitope, the virus often escapes via
nonsynonymous mutations that impair recognition. Synonymous mutations do not
affect this interplay and are often assumed to be neutral. We test this
assumption by tracking  synonymous mutations in 
longitudinal intrapatient data from the \shankaregion{} part of the \env{}
gene. We find that most synonymous variants are lost even though
they often reach high frequencies in the viral population, suggesting a
cost to the virus. Using published data
from SHAPE assays, we find that synonymous mutations that disrupt base pairs
in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
synonymous changes: these RNA hairpins might be important for HIV-1.
Computational modeling indicates that (i) most synonymous mutations in this
genomic region have a selection coefficient of the order of $-0.002$, and
(ii) synonymous variants rise to high frequency by genetic hitchhiking on neighboring beneficial
variants. This weak selection against synonymous substitutions does not
result in a strong pattern of conservation in cross sectional data, but
slows down the rate of evolution considerably. Our findings are consistent with the
notion that large scale patterns of RNA structure are functionally
relevant, while the precise base pairing pattern is not.
%\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using FFPopSim
\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
simulations where each site in the genome is represented by one bit that can be
in one of two states. Outcrossing rates, crossover rates, mutations rates and
arbitrary fitness functions can be specified. We used a generation time of 1
day, an outcrossing rate of $r=0.01$ per day \citep{batorsky_estimate_2011,
neher_recombination_2010}, a mutation rate of $\mu=10^{-5}$
\citep{mansky_lower_1995, abram_nature_2010} and simulated intrapatient
evolution for 6000 days. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
Mutations at the first and second positions were assigned strongly deleterious 
fitness effects 0.02. At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. Both full-length HIV-1 genomes and \env{}-only simulations
were performed and yielded comparable results.

The simulations were repeated 2400 times with random choices for the following
parameters: the fraction of deleterious sites $\alpha$ was sampled uniformly
between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
its logarithm was uniformly distributed  between $10^{-4}$ and $10^{-2}$; the
average escape rate $\epsilon$ of escape mutation was sampled such that its logarithm was
uniform between $10^{-2.5}$ and $10^{-1.5}$ and the rate $k_A$ of new antibody
challenges such that its logarithm was uniform between $10^{-3}$ and $10^{-2}$
per generation. Populations were initialized with a homogeneous founder
population and were kept at an average size of $N=10^4$ throughout the
simulation. After 30 generations of burn-in to create genetic diversity, new
epitopes were introduced at a constant rate $k_A$.

@@

@@

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, which shows the distribution of times after which synonymous
SNVs at intermediate frequencies become fixed or lost. The typical time to
loss is of the order of 500 days. If this loss is driven by the deleterious
effect of the mutation, this corresponds to deleterious effects $s_d$ of the
order of $- 0.002$ per day. (This is only an average estimate: every single
mutation is expected to have a slightly different fitness effect.)

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the
parameters, see \FIG{simsfig}. Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
0.002$. This is consistent with the fixation/extinction times estimated above
(see \FIG{fixp1}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4. We suggest through computational modeling that these
SNVs have deleterious effects on the order of $0.002$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.3\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.3\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
SNVs fixed (red), sites at which SNVs reached frequencies above 15\% but
were subsequently lost (blue), and sites at which no high-frequency SNVs were observed (green) 
(all categories are restricted to the regions V1-V5$\pm 100$bp).
Sites at which SNVs fixed tend to have higher SHAPE reactivities, corresponding to
less base pairing, than those at which SNVs are lost.
Sites at which no SNVs are observed show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous SNVs in
\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}
\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of synonymous SNVs. This parameter also reduces synonymous
diversity, measured by the probability of a SNV to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
day). Only simulations that reproduce the synonymous diversity and fixation
patterns observed in data are shown. These simulations demonstrate that
deleterious effects are around $-0.002$ and a large fraction of the 
synonymous mutations needs to be deleterious. As expected, larger
$s_d$ require larger $\epsilon$. Parameters are chosen
from prior distributions uniform in logspace as indicated by the red rectangle
(see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\pol}{\textit{pol}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\syndiv}{2}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
\newcommand{\gentimetwo}{5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Quantifying selection against synonymous mutations in HIV-1 \env{} evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{hyperref}
\hypersetup{colorlinks,linkcolor=red,citecolor=blue, pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\linespread{1.5}
\begin{document}
\title{\Title}
\author{\Author\\ \Affiliation}
%\affiliation{\Affiliation}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle

\newpage
%\begin{abstract}
\section*{Abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
or neutralizing antibodies target a new epitope, the virus often escapes via
nonsynonymous mutations that impair recognition. Synonymous mutations do not
affect this interplay and are often assumed to be neutral. We test this
assumption by tracking  synonymous mutations in 
longitudinal intrapatient data from the \shankaregion{} part of the \env{}
gene. We find that most synonymous variants are lost even though
they often reach high frequencies in the viral population, suggesting a
cost to the virus. Using published data
from SHAPE assays, we find that synonymous mutations that disrupt base pairs
in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
synonymous changes: these RNA hairpins might be important for HIV-1.
Computational modeling indicates that (i) most synonymous mutations in this
genomic region have a selection coefficient of the order of $-0.004$, and
(ii) synonymous variants rise to high frequency by genetic hitchhiking on neighboring beneficial
variants. This weak selection against synonymous substitutions does not
result in a strong pattern of conservation in cross sectional data, but
slows down the rate of evolution considerably. Our findings are consistent with the
notion that large scale patterns of RNA structure are functionally
relevant, while the precise base pairing pattern is not.
%\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using FFPopSim
\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
simulations where each site in the genome is represented by one bit that can be
in one of two states. Outcrossing rates, crossover rates, mutations rates and
arbitrary fitness functions can be specified. We used an outcrossing rate of
$r=0.01$ per day \citep{batorsky_estimate_2011, neher_recombination_2010},
a mutation rate of $\mu=2 \times 10^{-5}$ \citep{mansky_lower_1995, abram_nature_2010}.
We performed the simulations with
generation times of both 1 and 2 days, because this parameter varies across patients
\citep{perelson_hiv-1_1996, perelson_decay_1997}; our conclusions are shown in
did not change (see \FIG{simheat} and \figurename~S\gentimetwo, respectively). We simulated intrapatient
evolution for 6000 days. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
Mutations at the first and second positions were assigned strongly deleterious 
fitness effects 0.02. At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. Both full-length HIV-1 genomes and \env{}-only simulations
were performed and yielded comparable results.

The simulations were repeated 2400 times with random choices for the following
parameters: the fraction of deleterious sites $\alpha$ was sampled uniformly
between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
its logarithm was uniformly distributed  between $2 \cdot 10^{-4}$ and $2 \cdot 10^{-2}$; the
average escape rate $\epsilon$ of escape mutation was sampled such that its logarithm was
uniform between $2 \cdot 10^{-2.5}$ and $2 \cdot 10^{-1.5}$ and the rate $k_A$ of new antibody
challenges such that its logarithm was uniform between $10^{-3}$ and $10^{-2}$
per generation. Populations were initialized with a homogeneous founder
population and were kept at an average size of $N=10^4$ throughout the
simulation. After 30 generations of burn-in to create genetic diversity, new
epitopes were introduced at a constant rate $k_A$.

@@

% TODO change mu and rho and N

@@

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, which shows the distribution of times after which synonymous
SNVs at intermediate frequencies become fixed or lost. The typical time to
loss is of the order of 500 days. If this loss is driven by the deleterious
effect of the mutation, this corresponds to deleterious effects $s_d$ of the
order of $-0.002$ per day. (This is only an average estimate: every single
mutation is expected to have a slightly different fitness effect.)

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the
parameters, see \FIG{simsfig}. Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
0.004$. The inverse of $|s_d|$ is of the same order of magnitude as the
fixation/extinction times estimated above (see \FIG{fixp1}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4. We suggest through computational modeling that these
SNVs have deleterious effects on the order of $0.004$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.3\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.3\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
SNVs fixed (red), sites at which SNVs reached frequencies above 15\% but
were subsequently lost (blue), and sites at which no high-frequency SNVs were observed (green) 
(all categories are restricted to the regions V1-V5$\pm 100$bp).
Sites at which SNVs fixed tend to have higher SHAPE reactivities, corresponding to
less base pairing, than those at which SNVs are lost.
Sites at which no SNVs are observed show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous SNVs in
\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}
\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of synonymous SNVs. This parameter also reduces synonymous
diversity, measured by the probability of a SNV to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
day). Only simulations that reproduce the synonymous diversity and fixation
patterns observed in data are shown. These simulations demonstrate that
deleterious effects are around $-0.004$ and a large fraction of the 
synonymous mutations needs to be deleterious. As expected, larger
$s_d$ require larger $\epsilon$. Parameters are chosen
from prior distributions uniform in logspace as indicated by the red rectangle
(see methods).}
\label{fig:simheat}
\end{center}
\end{figure}
